A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-28 paper.tex
Editing pass

There are several techniques that aim to mitigate this complexity by
representing states symbolically.  Historically the most successful technique
has been to use \emph{Binary Decision Diagrams} (BDDs).  BDDs efficiently
represent a relation on a set of game variables but in the worst case the
representation may be exponential in the number of variables. This means that
BDDs are not a one-size-fits-all solution for all reactive synthesis
specifications.

Advances in SAT solving technology has prompted research into its applicability
to synthesis as an alternative to BDDs. One approach is to find sets of states
in CNF \cite{bloem2014,morgenstern2013}. Another approach is to eschew states
and focus on \emph{runs} of the game. Previous work has applied this idea to
realizability of bounded games \cite{narodytska2014} by forming abstractions of
the game and refining in a counterexample-guided framework. This has been shown
to outperform BDDs on certain classes of game specifications but is only able
to find counterexamples of a certain length to the safety property.  In this
paper, we extend this idea to unbounded games by approximating sets of unsafe
states from abstract games. Careful construction ensures that a fixed point in
these sets guarantees completeness.

@@

Realisability for a bounded safety game can be solved by checking an
quantified formula consisting of the game's transition relation \emph{unrolled}
to the length of the bounded game. The quantifiers represent the two players of
the game: universal for the controller and existential for the environment.
Just as the players alternate in choosing actions every step of the game, there
must be quantifier alternations in the formula for each unrolling.

Instead of solving both quantifiers at once, two competing solvers are used to
solve abstractions of the game. Each solver builds a candidate strategy for its
corresponding player and uses its opponent's candidate to restrict its own
search. The solvers communicate to refine one another's candidate strategies by
finding counterexamples. Eventually one solver will be unable to find a
counterexample indicating that its opponent's candidate strategy is a winning
strategy for the game.

@@

Given a controller (environment)
abstract game tree $T$ a \emph{partial strategy} $Strat: \textsc{nodes}(T) \rightarrow L_u$ 
($Strat: \textsc{nodes}(T) \rightarrow L_c$) labels each node of the tree with the opponent's
action to be played in this node.   
Figure~\ref{fig:strategy} shows an example partial strategy.  The environment 
starts by choosing action $\alpha$.  If the controller plays $a$, the
environment responds with $\beta$ in the next round, and so on. Given a partial
strategy $Strat$, we can map each leaf $l$ of the abstract game tree to
$\langle s',i'\rangle=\textsc{outcome}(\langle s, i\rangle, Strat, l)$ obtained
by playing all controllable and uncontrollable actions on the path from the
root to the leaf.

@@

The algorithm takes a concrete game $G$ with bound $n$ as an implicit argument.
In addition, it takes a player $p$ (controller or environment), state $s$,
bound $k$ and an abstract game tree $T$ and returns a winning partial strategy
for $opponent(p)$, if one exists.  The initial invocation of the algorithm
takes the initial state $\langle I, n\rangle$ and an empty abstract game tree
$\emptyset$.  Initially the solver is playing on behalf of the environment
since that player takes the first move in every step.  The empty game tree does
not constrain opponent moves, hence solving such an abstraction is equivalent
to solving the original concrete game.

@@

The way game trees are refined is determined by the formula passed to the SAT
solver in $\textsc{findCandidate}$. This formula contains CNF encodings of all
of the unrolled runs represented by an abstract game tree and the winning
condition of the current player. Runs are encoded by copying the transition
relation for every step in the abstract game. When playing for the controller,
the SAT solver searches for a satisfying assignment to the unfixed label
variables in tree so that none of the runs reaches the error state. The
environment formulation is satisfiable if any run does reach the error state.
The formula is constructed recursively from the root of a tree by
$\textsc{treeFormula}$ (see Algorithm~\ref{alg:treeFormula}).

Since the game tree formulation is passed to a SAT solver, both controllable
and uncontrollable unfixed labels will be existentially quantified. This means
that the SAT solver will find any way to win the game while both players are
cooperating. If no winning run exists in an abstract game even when the players
are cooperating then there is no winning run when the opponent is playing
adversarily. When a winning run is found, the actions chosen by the SAT solver
are used to refine the game tree. This is advantageous for many synthesis
problems where the game must be formalised as adversarial for correctness but
the final implementation will be cooperating with its environment in the real
world. An example of such a system is a device driver that cooperates with the
device and OS to interface between the two.

@@

A similar concept can be applied to the bounded synthesis algorithm to
iteratively increase the bound of the game and terminate when a fixed point is
reached. When a strategy is found to be winning on an abstract game tree, we
record as winning the states from which the opponent could find no
counterexample. To find these states we use interpolation of subformulas of the
game tree.

@@

The states in an abstract game with no controller candidate are
\emph{must-losing}. The controller will never be able to avoid the error state
from these states.

From abstractions with no environment candidate we record the complement states
as \emph{may-losing}. We maintain a set of states for every rank up the current
bound. The states we learn while executing the bounded algorithm can only said
to be controller-winning for a particular rank or below. However, we construct
each set carefully such that it is monotonically larger than the set of the
rank below
%(see Figure \ref{fig:maylosing})
and so that the environment is unable to force play from one set to the next.
Due to this careful construction, when two adjacent sets become equivalent we
know that the algorithm has reached a fixed point and the controller is winning
in the unbounded game (line 6).

@@

Given two formulas $A$ and $B$ such that $A \land B$ is unsatisfiable, it is
possible to construct a Craig interpolant\cite{craig1957} $\mathcal{I}$ such
that $A \to \mathcal{I}$, $B \land \mathcal{I}$ is unsatisfiable, and
$\mathcal{I}$ refers only to the intersection of variables in $A$ and $B$.  An
interpolant can be constructed efficiently from a resolution proof of the
unsatisfiability of $A \land B$ \cite{pudlak1997}.

@@

This is the foundation for a recursive algorithm that consumes an entire tree
by removing a single step on each iteration (Algorithm \ref{alg:learn}). All
learned states from which the controller \emph{must} lose are recorded in a set
of \emph{bad} states $B^M$.  This algorithm can also be performed on
environment-losing trees with the caveat that any state learnt at a node of
rank $n$ is only known to lose at ranks less than or equal to $n$. We negate
the interpolants we learn this way and record them in a mapping of ranks to
sets of states $B^m$ which \emph{may} be bad states for the controller.

@@

Line~(11) splits the tree $T$ into $T_1$ and $T_2$, such that $T_2$ has depth
1.  Consider formulas $f_1=\sigma(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ and $F_2 =
\textsc{\textoverline{treeFormula}}(T_2)$.  These formulas only share variables
$X_n$.  Their conjunction $F_1 \land F_2$ is unsatisfiable, as by construction
any solution of $F_1 \land F_2$ also satisfies $\sigma(X_T) \land
\textsc{treeFormula}(T)$, which is unsatisfiable (precondition (b)).  Hence the
interpolation operation is defined for $F_1$ and $F_2$.

@@

\subsection{Choosing sensible opponent moves}

>>>>

There are several techniques that aim to mitigate this complexity by
representing states symbolically.  Historically the most successful technique
has been to use \emph{Binary Decision Diagrams}.  BDDs efficiently
represent a relation on a set of game variables but in the worst case the
representation may be exponential in the number of variables. As a result
BDDs are not a one-size-fits-all solution for all reactive synthesis
specifications.

Advances in SAT solving technology has prompted research into its applicability
to synthesis as an alternative to BDDs. One approach is to find sets of states
in CNF \cite{bloem2014,morgenstern2013}. Another approach is to eschew states
and focus on \emph{runs} of the game. Previous work has applied this idea to
realizability of bounded games \cite{narodytska2014} by forming abstractions of
the game and refining in a counterexample-guided framework. This has been shown
to outperform BDDs on certain classes of game specifications but is only able
to find counterexamples to the safety property of a certain length.  In this
paper, we extend this idea to unbounded games by approximating sets of unsafe
states from abstract games. Careful construction ensures that a fixed point in
these sets guarantees completeness.

@@

Realisability for a bounded safety game can be solved by checking a quantified
formula consisting of the game's transition relation \emph{unrolled} to the
length of the bounded game. The quantifiers represent the two players of the
game: universal for the controller and existential for the environment.  Just
as the players alternate in choosing actions every step of the game, the
formula is prefixed by a quantifier alternation over $L_u$ and $L_c$ for each
step.

Instead of solving both quantifiers at once, two competing solvers are used to
solve abstractions of the game. Each solver builds a candidate strategy for its
corresponding player and uses its opponent's candidate to restrict its own
search. The solvers communicate to refine one another's candidate strategies by
finding counterexamples. Eventually one solver will be unable to find a
counterexample, indicating that its opponent's candidate strategy is a winning
strategy for the game.

@@

Given a controller (environment)
abstract game tree $T$ a \emph{partial strategy} $Strat: \textsc{nodes}(T) \rightarrow L_u$ 
($Strat: \textsc{nodes}(T) \rightarrow L_c$) labels each node of the tree with
the opponent's action to be played in that node.   Figure~\ref{fig:strategy}
shows an example partial strategy.  The environment starts by choosing action
$\alpha$.  If the controller plays $a$, the environment responds with $\beta$
in the next round, and so on. Given a partial strategy $Strat$, we can map each
leaf $l$ of the abstract game tree to $\langle
s',i'\rangle=\textsc{outcome}(\langle s, i\rangle, Strat, l)$ obtained by
playing all controllable and uncontrollable actions on the path from the root
to the leaf.

@@

The algorithm takes a concrete game $G$ with maximum bound $n$ as an implicit
argument.  In addition, it takes a player $p$ (controller or environment),
state $s$, bound $k$ and an abstract game tree $T$ and returns a winning
partial strategy for $\textsc{opponent}(p)$, if one exists.  The initial
invocation of the algorithm takes the initial state $\langle I, n\rangle$ and
an empty abstract game tree $\emptyset$.  Initially the solver is playing on
behalf of the environment since that player takes the first move in every step.
The empty game tree does not constrain opponent moves, hence solving such an
abstraction is equivalent to solving the original concrete game.

@@

Candidates are discovered by passing a formulation of the abstract game tree to
a SAT solver in $\textsc{findCandidate}$. This formula contains CNF encodings
of all of the unrolled runs represented by the tree and the winning condition
of the current player.  Runs are encoded by copying the transition relation for
every step in the abstract game. When playing for the controller, the SAT
solver searches for a satisfying assignment to the unfixed label variables in
tree so that none of the runs reaches the error state. The environment
formulation is satisfiable if any run does reach the error state.  The formula
is constructed recursively from the root of a tree by $\textsc{treeFormula}$
(see Algorithm~\ref{alg:treeFormula}).

Since the game tree formulation is passed to a SAT solver, both controllable
and uncontrollable unfixed labels will be existentially quantified. This means
that the SAT solver will find any way to win the game while both players are
cooperating. If no winning run exists in an abstract game even when the players
are cooperating then there is no winning run when the opponent is playing
adversarily. When a winning run is found, the actions chosen by the SAT solver
are used to refine the game tree. This is advantageous for many synthesis
problems where the game must be formalised as adversarial for correctness but
the final implementation will cooperate with its environment in the real world.
An example of such a system is a device driver that cooperates with the device
and OS to provide the interface between the two.

@@

We apply a similar fixed point computation on game states to the bounded
synthesis algorithm in order to promote it to unbounded games. Our unbounded
procedure iteratively executes the bounded algorithm while recording states
from abstract game trees. By carefully managing the way states are recorded we
guarantee that a fixed point proves the controller's strategy correct.

@@

The states in an abstract game with no controller candidate are
\emph{must-losing}. The environment can always force the game into the error
states from these states.

From abstractions with no environment candidate we record the complement of
states in the tree as \emph{may-losing}. The environment cannot reach the error
state in a number of steps equal to the distance to the bottom of the tree.

We maintain a set of states for each rank up to the current bound. We maintain
an invariant over these sets via careful construction so that they are
monotonically increasing by rank. We also ensure that the environment is unable
to force play from one set to the next. Due to these invariants, when two
adjacent sets become equivalent we know that the algorithm has reached a fixed
point and the controller is winning in the unbounded game (line 6).

@@

Given two formulas $F_1$ and $F_2$ such that $F_1 \land F_2$ is unsatisfiable,
it is possible to construct a Craig interpolant\cite{craig1957} $\mathcal{I}$
such that $F_1 \to \mathcal{I}$, $F_2 \land \mathcal{I}$ is unsatisfiable, and
$\mathcal{I}$ refers only to the intersection of variables in $F_1$ and $F_2$.
An interpolant can be constructed efficiently from a resolution proof of the
unsatisfiability of $F_1 \land F_2$ \cite{pudlak1997}.

@@

This is the foundation for a recursive algorithm that consumes an entire tree
by removing a single step on each iteration (Algorithm \ref{alg:learn}). All
learned states from which the controller must lose are recorded in a set of bad
states $B^M$. Environment-losing are learned in the same way but are recorded
by removing the state from a set of sets of controller may-losing states $B^m$.
An environment-losing state of rank $n$ is removed from every set at rank $n$
and lower.

@@

Line~(11) splits the tree $T$ into $T_1$ and $T_2$, such that $T_2$ has depth
1.  Consider formulas $F_1=\sigma(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ and $F_2 =
\textsc{\textoverline{treeFormula}}(T_2)$.  These formulas only share variables
$X_n$.  Their conjunction $F_1 \land F_2$ is unsatisfiable, as by construction
any solution of $F_1 \land F_2$ also satisfies $\sigma(X_T) \land
\textsc{treeFormula}(T)$, which is unsatisfiable (precondition (b)).  Hence the
interpolation operation is defined for $F_1$ and $F_2$.

@@

\subsection{Choosing opponent moves}
